---
title: "King Co. Synthesis - Metia"
author: "Anders Swanson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    depth: 3
    toc: yes
    theme: spacelab
---
#1.	Define Project Scope

##The Task

> Using whatever tools you have at your disposal (though ideally Power BI) we would like you to find two datasets that you can integrate (i.e. they have a meaningful primary key in common) at the site below. https://data.kingcounty.gov/browse

> We’d like you to integrate, analyze, define insights, and visualize the output. No more than 2-4 hours of time should go into this – we’re looking for insight into your ‘quick take’ on data and storytelling rather than a polished solution.

Based on the description of the task above, I divided the project into the six steps listed below.

## Outline of Work
1.	Define Project Scope (15 min)
2.	Acquire Data (30 min)
3.	Merge (& Clean) Data (45 min)
4.	Explore Synthesized Data for Insights (45 min)
5.	Visualize Insights  (30)
6.  Evaluation (15 min)
7.  Share Findings (30)

#2.	Acquire Data

##Explore Data.KingCounty.gov

I found a few things of interest
#3.	Merge Data
#4.	Explore Synthesized Data for Insights
#5.	Share Insights & Findings
#6. Evaluation

```{r load packages, message =  FALSE}
#load necessary packages
library(lubridate)
library(dplyr)
```

Let's start by loading into memory the [Food Establishment Inspection Database](https://data.kingcounty.gov/Health/Food-Establishment-Inspection-Data/f29f-zza5).


```{r get data}

#download data Data directory read into memory
download.file("https://data.kingcounty.gov/api/views/f29f-zza5/rows.csv", "./data/Food_Establishment_Inspection_Data.csv")

food <-read.csv("data/Food_Establishment_Inspection_Data.csv")

#create random sample of 1000 inspections
food.sample <- food %>% sample_n(1000)
```




